---
title: "Homework 5"
author: "Sricharraan Ramaswamy"
date: "09/03/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
load("lattice.Rdata")
head(hamster)
```

```{r}
library(devtools)
install_github("vqv/ggbiplot")
```

```{r}
library(ggbiplot)
library(ade4)
library(viridis)
```

```{r}
summary(hamster)
ham_scale_1 = prcomp(hamster , scale. = FALSE)
ham_scale_1$rotation[,1:2]
ham_scale_2 = prcomp(hamster , scale. = TRUE)
ham_scale_2$rotation[,1:2]
ggbiplot(ham_scale_1,scale = 0) + ggtitle("Without scaling")
ggbiplot(ham_scale_2,scale = 0) + ggtitle("With scaling")
```
The principal-axes capture just the variance of the features 'Liver' and 'Testes,' which have the most variation in the data set, when PCA is applied to unscaled log-transformed weights. These two axis have relatively big eigenvalues. 
The biplot's other principal-axes (designated as combinations of organ-weights) appear congested.

The eigenvalues for 'Liver' and 'Testes' no longer dominate the principal axes thanks to feature-scaling. 
In addition, the data appears to be projected more accurately onto the dimensions with great variance.

Therefore, in this instance the one which is scaled is giving more accurate results than the one without scaling 
```{r}
word_freq <- read.csv("http://jfukuyama.github.io/teaching/stat670/assignments/nyt_articles.csv")
```

```{r}
unique(word_freq$class.labels)
```

```{r}
which(apply(word_freq, 2, var)==0)
```

```{r}
drops <- c("of","the","art")
word_freq <- word_freq[ , !(names(word_freq) %in% drops)]
```


```{r}
word_freq_1 = prcomp(word_freq[-1] , scale. = FALSE, center = TRUE)
word_freq_2 = prcomp(word_freq[-1] , scale. = TRUE, center = TRUE)
df1 = data.frame(word_freq_1$x)
df1$class.labels <- word_freq$class.labels
df2 = data.frame(word_freq_2$x)
df2$class.labels <- word_freq$class.labels

ggbiplot(word_freq_1, scale = 1, var.axes = FALSE, groups = df1$class.labels) + ggtitle("without Scaling")
ggbiplot(word_freq_2, scale = 1, var.axes = FALSE, groups = df2$class.labels) + ggtitle("with Scaling")
  
```
 
The scatterplot of points projected on the principal axes demonstrates a substantial separation between 'art' and'music' articles when the data is not scaled. 
The eigenvalue of PC-2 of'music' articles is tiny, but PC-2 of 'art' articles is huge. 
When the dataset is scaled, the result is a more congested view.

```{r}

word_freq_1 = prcomp(word_freq[-1] , scale. = FALSE, center = TRUE)

n_loadings <- 10
loadings <- data.frame(word_freq_1$rotation[,1:2])
loadings_0 <- loadings[order(-loadings$PC1),][1:n_loadings,]
loadings_1 <- loadings[order(-loadings$PC2),][1:n_loadings,]

ggplot() + geom_point(aes(x=PC1, y=PC2, color=class.labels), data = df1) + geom_segment(data=loadings_0,aes(xend=PC1*3, yend=PC2*3, x=0, y=0)) + geom_text(data=loadings_0, aes(x=PC1*3, y=PC2*3, label=row.names(loadings_0))) + xlab("PC 1") + ylab("PC 2") + ggtitle(" 10 Loadings ordered by PC 1") + scale_color_discrete('Class labels')

ggplot() + geom_point(aes(x=PC1, y=PC2, color=class.labels), data = df1) + geom_segment(data=loadings_1,aes(xend=PC1*3, yend=PC2*3, x=0, y=0)) + geom_text(data=loadings_1, aes(x=PC1*3, y=PC2*3, label=row.names(loadings_0))) + xlab("PC 1") + ylab("PC 2") + ggtitle(" 10 Loadings ordered by PC 2") + scale_color_discrete('Class labels')
```
The words classified as'music' articles contribute more to the largest loadings on PC 1. As a result, PC 1 now controls the majority of the 'music' articles. 

The terms shown as 'art' articles contribute the most to the highest loadings on PC 2. This means that PC 2 takes up the majority of the 'art' articles' variation. 

Finally, we can see that each Principal component only receives variance from one class label, such as 'art' or'music.' As a result, the art articles share extremely few words with the music entries.